### 2023-07-12 submission/manuscript.tex
Submitted to journal
Use gemelli to perform rclr/rpca

The practice of rarefaction has been commonly used in ecology for more
than 50 years as a tool to control for uneven sequencing effort across
experimental replicates (4, 5). Microbial ecologists have used it to
compare 16S rRNA gene sequence data for the past 25 years (6--8). With
rarefaction the investigator selects a desired threshold of sequencing
effort and removes any samples below that threshold. They then randomly
select that many sequences, with replacement from each sample. Based on
the observed sequence counts, the researcher can then calculate alpha
diversity metrics including richness and diversity indices or beta
diversity metrics such as a Jaccard or Bray-Curtis dissimilarity index.
I refer to this single sampling as a subsample; this method is
implemented as the \texttt{sub.sample} function in mothur (9) and the
\texttt{rrarefy} function in the \texttt{vegan} R package (10).
Rarefaction repeats the subsampling a large number of times (e.g., 100
or 1,000 times) and calculates the mean of the alpha or beta diversity
metric over those subsamplings; rarefaction is implemented in mothur
using the \texttt{summary.single} and \texttt{dist.shared} functions (9)
and with the \texttt{vegan} R package using the \texttt{rarefy} or
\texttt{avgdist} functions (10). Rarefaction effectively tells a
researcher what an alpha or beta diversity metric would have been for a
collection of samples if they were all sequenced to the same depth.
Although a closed form equation exists to calculate the expected
richness (5), it is computationally easier to empirically calculate
richness and other alpha and beta diversity metrics by rarefaction.

@@

\textbf{\emph{Without rarefaction, metrics of beta diversity are
sensitive to sequencing effort.}} To test the sensitivity of various
approaches of measuring beta diversity to sequencing effort, I used the
same null models used for studying the sensitivity of alpha diversity.
Under a null model, the ecological distance between any pair of samples
would be the same regardless of the difference in the number of
sequences observed in each sample (e.g., Figure S3). First, I calculated
the Jaccard distance coefficient between all pairs of communities within
a dataset. The Jaccard distance coefficient is the fraction of OTUs that
are unique to either community and does not account for the abundance of
the OTUs. Jaccard distances were calculated using the uncorrected OTU
counts, with rarefaction, relative abundances, and following
normalization using cumulative sum scaling (CSS) and SRS. Only the
distances calculated with rarefaction showed a lack of sensitivity to
sequencing effort (Figure 1B). Second, I analyzed the sensitivity of the
Bray-Curtis distance coefficient, which is a popular metric that
incorporates the abundance of each OTU. Similar to what I observed with
the Jaccard coefficient, only the data normalized with rarefaction
showed a lack of sensitivity to sequencing effort (Figure 1B). Third, I
calculated the Euclidean distance on raw OTU counts where the central
log-ratio (CLR) was calculated (i.e., Aitchison distances) by ignoring
OTUs in samples with zero counts (Robust CLR), adding a pseudo-count of
one to all OTU counts prior to calculating the CLR (One CLR), adding a
pseudo-count of one divided by the total number of sequences obtained
for the community (Nudge CLR), and imputing the value of zero counts
(Zero CLR). The Aitchison distances were all strongly sensitive to
sequencing effort (Figure 1B). Finally, I used the
variance-stabilization transformation (VST) from DESeq2 prior to
calculating Euclidean distances. Again, there was a strong sensitivity
to sequencing effort (Figure 1B). Although Euclidean distances are not
typically used on raw or rarefaction-normalized count data in ecology
(26), Euclidean distances calculated with rarefaction were not sensitive
to sequencing effort. Across each of the beta diversity metrics and
approaches used to account for uneven sequencing effort and sparsity,
rarefaction was the least sensitive approach to differences in
sequencing effort.

@@

\textbf{Richness-adjusted community model.} In the richness-adjusted
community model, communities were randomly assigned to one of two
simulated treatment groups. Communities in the first treatment group
were generated by calculating the relative abundance of each OTU across
all samples and using those values as the probability of sampling each
OTU. This probability distribution was sampled until each sample had the
same number of sequences that it did in the observed data. Samples in
the second treatment group were generated by removing 3\% of the OTUs
from the dataset and recalculating the relative abundance of the
remaining OTUs. Sequences were sampled from the richness-adjusted
community distribtion, with replacement, until each sample had the same
number of sequences that it did in the observed data. Under the
richness-adjusted community model, each sample represented a statistical
sampling of two communities such that there should not have been a
statistically significant difference within a treatment group, but there
was between the treatment groups. Because the construction of the
richness-adjusted community model was a stochastic process, 100
replicates were generated for each dataset.

@@

where \(x_{ij}\) was the number of sequences observed for OTU \(i\) in
sample \(j\) and \(g()\) was the geometric mean, \(x_j\) was the count
of the \(S_T\) OTUs in sample \(j\). Because the geometric mean is zero
if any OTU is absent from a sample, the CLR is undefined when there are
unobserved OTUs in a sample. To overcome this problem, I attempted a
four approaches. The first, Zero CLR, imputed the value of the zeroes
based on the observed data using the zCompositions (v.1.4.0.1) R package
(34). The second, One CLR, added a pseudo-count of 1 to the abundance of
all OTUs (18, 20). The third, Nudge CLR, added a pseudo-count of 1
divided by the total number of sequences in a sample to each OTU in the
sample (18, 23). The final approach, Robust CLR, removed unobserved OTUs
prior to calculating the CLR (22).

@@

\leavevmode\vadjust pre{\hypertarget{ref-Schloss2023wnwn}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{\textbf{Schloss PD}. 2023. Waste not, want not:
Revisiting the analysis that called into question the practice of
rarefaction. bio{R}xiv.
doi:\href{https://doi.org/Pending\%20at\%20bioRxiv}{Pending at
bioRxiv}.}

>>>>

The practice of rarefaction has been commonly used in ecology for more
than 50 years as a tool to control for uneven sequencing effort across
experimental replicates (4, 5). Microbial ecologists have used it to
compare 16S rRNA gene sequence data for the past 25 years (6--8). With
rarefaction the investigator selects a desired threshold of sequencing
effort and removes any samples below that threshold. They then randomly
select that many sequences, without replacement from each sample. Based
on the observed sequence counts, the researcher can then calculate alpha
diversity metrics including richness and diversity indices or beta
diversity metrics such as a Jaccard or Bray-Curtis dissimilarity index.
I refer to this single sampling as a subsample; this method is
implemented as the \texttt{sub.sample} function in mothur (9) and the
\texttt{rrarefy} function in the \texttt{vegan} R package (10).
Rarefaction repeats the subsampling a large number of times (e.g., 100
or 1,000 times) and calculates the mean of the alpha or beta diversity
metric over those subsamplings; rarefaction is implemented in mothur
using the \texttt{summary.single} and \texttt{dist.shared} functions (9)
and with the \texttt{vegan} R package using the \texttt{rarefy} or
\texttt{avgdist} functions (10). Rarefaction effectively tells a
researcher what an alpha or beta diversity metric would have been for a
collection of samples if they were all sequenced to the same depth.
Although a closed form equation exists to calculate the expected
richness (5), it is computationally easier to empirically calculate
richness and other alpha and beta diversity metrics by rarefaction.

@@

\textbf{\emph{Without rarefaction, metrics of beta diversity are
sensitive to sequencing effort.}} To test the sensitivity of various
approaches of measuring beta diversity to sequencing effort, I used the
same null models used for studying the sensitivity of alpha diversity.
Under a null model, the ecological distance between any pair of samples
would be the same regardless of the difference in the number of
sequences observed in each sample (e.g., Figure S3). First, I calculated
the Jaccard distance coefficient between all pairs of communities within
a dataset. The Jaccard distance coefficient is the fraction of OTUs that
are unique to either community and does not account for the abundance of
the OTUs. Jaccard distances were calculated using the uncorrected OTU
counts, with rarefaction, relative abundances, and following
normalization using cumulative sum scaling (CSS) and SRS. Only the
distances calculated with rarefaction showed a lack of sensitivity to
sequencing effort (Figure 1B). Second, I analyzed the sensitivity of the
Bray-Curtis distance coefficient, which is a popular metric that
incorporates the abundance of each OTU. Similar to what I observed with
the Jaccard coefficient, only the data normalized with rarefaction
showed a lack of sensitivity to sequencing effort (Figure 1B). Third, I
calculated the Euclidean distance on raw OTU counts where the central
log-ratio (CLR) was calculated (i.e., Aitchison distances) using matrix
completion to replace zero counts (Robust CLR), adding a pseudo-count of
one to all OTU counts prior to calculating the CLR (One CLR), adding a
pseudo-count of one divided by the total number of sequences obtained
for the community (Nudge CLR), and imputing the value of zero counts
(Zero CLR). The Aitchison distances were all strongly sensitive to
sequencing effort (Figure 1B). Finally, I used the
variance-stabilization transformation (VST) from DESeq2 prior to
calculating Euclidean distances. Again, there was a strong sensitivity
to sequencing effort (Figure 1B). Although Euclidean distances are not
typically used on raw or rarefaction-normalized count data in ecology
(26), Euclidean distances calculated with rarefaction were not sensitive
to sequencing effort. Across each of the beta diversity metrics and
approaches used to account for uneven sequencing effort and sparsity,
rarefaction was the least sensitive approach to differences in
sequencing effort.

@@

\textbf{Richness-adjusted community model.} In the richness-adjusted
community model, communities were randomly assigned to one of two
simulated treatment groups. Communities in the first treatment group
were generated by calculating the relative abundance of each OTU across
all samples and using those values as the probability of sampling each
OTU. This probability distribution was sampled until each sample had the
same number of sequences that it did in the observed data. Samples in
the second treatment group were generated by removing 3\% of the OTUs
from the dataset and recalculating the relative abundance of the
remaining OTUs. Sequences were sampled from the richness-adjusted
community distribution, with replacement, until each sample had the same
number of sequences that it did in the observed data. Under the
richness-adjusted community model, each sample represented a statistical
sampling of two communities such that there should not have been a
statistically significant difference within a treatment group, but there
was between the treatment groups. Because the construction of the
richness-adjusted community model was a stochastic process, 100
replicates were generated for each dataset.

@@

where \(x_{ij}\) was the number of sequences observed for OTU \(i\) in
sample \(j\) and \(g()\) was the geometric mean, \(x_j\) was the count
of the \(S_T\) OTUs in sample \(j\). Because the geometric mean is zero
if any OTU is absent from a sample, the CLR is undefined when there are
unobserved OTUs in a sample. To overcome this problem, I attempted a
four approaches. The first, Zero CLR, imputed the value of the zeroes
based on the observed data using the zCompositions (v.1.4.0.1) R package
(34). The second, One CLR, added a pseudo-count of 1 to the abundance of
all OTUs (18, 20). The third, Nudge CLR, added a pseudo-count of 1
divided by the total number of sequences in a sample to each OTU in the
sample (18, 23). The final approach, Robust CLR, used matrix completion
to replace zero counts for unobserved OTUs prior to calculating the CLR
(22).

@@

\leavevmode\vadjust pre{\hypertarget{ref-Schloss2023wnwn}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{\textbf{Schloss PD}. 2023. Waste not, want not:
Revisiting the analysis that called into question the practice of
rarefaction. bio{R}xiv.
doi:\href{https://doi.org/10.1101/2023.06.23.546312}{10.1101/2023.06.23.546312}.}
